---
title: "Intro to MAF"
author: Nicole Black, Wade Boohar
date: 07/17/22
updated: 09/17/23
---

***Deliverables***
-Upload this R Notebook to your GitHub and submit the link to your Repo on BB.
-Include any graphs or figures created in this assignment in the folder with your R notebook with descriptive file names.

Since this is an optional partner activity, it is okay if your answers are the same as your partner’s as long as everyone understands it and could explain it in their own words if asked. Each person must individually push their code to Github. *At the top of your R Notebook, write the name of you and your partner(s) as a comment.*

***Complete the following coding activity and answer any following questions as comments in your R Notebook***

In MAF_Tutorial.Rmd, you looked at how to analyze mutation data in terms of clinical variables as well as mutation status. In this assignment, you will need to combine the two skills to demonstrate your understanding of categorical variables and R data structures.

*Pre-Assignment*
Load all necessary packages, read in the clinical data .csv file you have in your analysis_data folder, and instantiate the MAF_object.
```{r}
MAF_object <- read.csv("C:/Users/josep/Desktop/QBIO490/qbio_490_joseph_kim/analysis_data/skcm_clinical_data.csv")
clinical <- maf_object@clinical.data

install.packages("tidyverse")
library(tidyverse)
```


*1*
Choose a clinical variable (or any variable from clin_rad or clin_drug) to separate your populations into two different groups and rewrite the column or create a new column with that variable as a factor. ***Do not use age or vital_status as your clinical variable.*** Hint: if your variable is continuous, you will need to determine your own cutoffs for the different levels of the factor. If your variable is categorical and has more than two possible values, choose the two that are the most common.
```{r}
na_mask <- ifelse(MAF_object$weight_kg_at_diagnosis == '[Not Available]', FALSE, TRUE)
MAF_object <- MAF_object[na_mask, ]
MAF_object$weight_kg_at_diagnosis <- as.numeric(MAF_object$weight_kg_at_diagnosis)

median(MAF_object$weight_kg_at_diagnosis)
MAF_object$clinVar <- ifelse(MAF_object$weight_kg_at_diagnosis >= 80, "HEAVY", "LIGHT")
#MAF_object$clinVar



na_mask_ <- ifelse(clinical$weight_kg_at_diagnosis == '[Not Available]', FALSE, TRUE)
clinical <- clinical[na_mask_, ]
clinical$weight_kg_at_diagnosis <- as.numeric(clinical$weight_kg_at_diagnosis)
clinical$cateWeight <- ifelse(clinical$weight_kg_at_diagnosis >= 80, "HEAVY", "LIGHT")
```


*2*
Create a co-oncoplot with the top 10-20 (you choose) most mutated genes for the two groups. Pick one that has a large discrepancy in % mutated or type of mutations between the groups and research it. 
Research it. What is the gene used for? Can you think of any reason for the discrepancy?
```{r}
heavy_barcodes <- clinical[clinical$cateWeight == "HEAVY", ]
heavy_barcodes <- heavy_barcodes$Tumor_Sample_Barcode

heavy_maf <- subsetMaf(maf = maf_object, tsb = heavy_barcodes)

light_barcodes <- clinical[clinical$cateWeight == "LIGHT", ]
light_barcodes <- light_barcodes$Tumor_Sample_Barcode

light_maf <- subsetMaf(maf = maf_object, tsb = light_barcodes)

#Researching with help and the internet, I could not find a function in coOncoplot that increases the number of genes that appear, so I stuck with the default. There is a function for normal oncoplots, but not for co-oncoplots.
coOncoplot(m1 = heavy_maf,
           m2 = light_maf,
           m1Name = "Heavy (>= 80kg)",
           m2Name = "Light (< 80kg)",
           borderCol = NA,
           )
```
There seems to be a discrepancy in the DNAH5 gene comparing those under and over 80kg in weight. The DNAH5 gene provides instructions for a protein in a group for dynein, which produces the force needed for a cell's cilia to move. This is closely linked to primary ciliary dyskinesia, a condition primarily linked with respiratory tract infections. Regarding weight, those with respiratory tract infections are likely to be heavier due to the issues regarding exercise. 

*3*
Create a contingency table with your variable and chosen gene. Run a Fisher’s Exact Test between presence of mutations for that gene and your clinical variable. Create and save a mosaic plot. 
Interpret the output of the Fisher’s Exact Test in terms of the odds ratio and p-value.
```{r}
barcodes <- clinical$Tumor_Sample_Barcode

DNAH5 <- subsetMaf(maf = maf_object, genes = "DNAH5", tsb = barcodes)
DNAH5@data <- DNAH5@data[!duplicated(DNAH5@data$Tumor_Sample_Barcode), ]

DNAH5@data$Variant_Classification <- factor(DNAH5@data$Variant_Classification)
DNAH5@clinical.data$cateWeight <- factor(DNAH5@clinical.data$cateWeight)

contig <- table(DNAH5@data$Variant_Classification, DNAH5@clinical.data$cateWeight)
contig

mosaicplot(contig)

fisher_test <- fisher.test(contig)
fisher_test

```
The Fisher's Exact Test between weight and DNAH5 mutations return a p-value of 0.6797: there is no significant association with this p-value under any alpha value. Therefore, we cannot conclude that there exists an association between weight and DNAH5 mutations.

With regards to the odd's table, the resulting odd's ratio between missense and nonsense mutations equals 0.89, showing a 0.89x chance of having a missense mutation if one is heavier than 80kg.

*4*
Subset your maf_object based on your chosen clinical variable and create a co-lollipop plot of your chosen gene divided between the two different clinical variable possibilities. Include descriptive names on your plot.
Do you notice any difference in terms of mutations (e.g. sites, types, number) between the two populations?
```{r}
light_maf <- subsetMaf(maf = maf_object, tsb = light_barcodes)
heavy_maf <- subsetMaf(maf = maf_object, tsb = heavy_barcodes)

lollipopPlot2(m1 = heavy_maf,
              m2 = light_maf,
              m1_name = "Heavy (>= 80kg)",
              m2_name = "Light (< 80kg)",
              gene = "DNAH5")
```
In terms of mutations, the are less mutations in those who are lighter than 80kg, which is something expected. The heavier group has more splice_site mutations, while both groups primarily have a similar amount of missense mutations. The biggest difference is the presence of mutations around the NTPase on the gene -- there are significantly more mutations on the heavier side than the lighter for that specific site. 

*5*
Create your Overall_Survival_Status column and create a mafSurvival KM plot based on mutations in your chosen gene.
Does there seem to be a difference? Hypothesize why or not based on the other analysis you did with the gene above.
```{r}
maf_object@clinical.data$Overall_Survival_Status <- ifelse(maf_object@clinical.data$vital_status == "Alive", TRUE, FALSE)
maf_object@clinical.data$Overall_Survival_Status

mafSurvival(maf = maf_object, genes = "DNAH5", time = "last_contact_days_to", Status = "Overall_Survival_Status", isTCGA = TRUE)
```
There seems to be a small difference in survival time between those heavier and lighter with mutations in DNAH5. However, the mutation is associated with a higher survival rate, which is not what is expected -- and since the P-value of this association is 0.216, higher than any significant alpha value, we cannot conclude an association between mutation and survival rate. As analyzed previously, there does not appear to be an association between weight and mutations in DNAH5.
I hypothesize that the reason there is a lack of association is due to the small dataset that I chose to analyze. DNAH5 is a minor gene part of a larger complex with little data both in the maf dataframe as well as online. Therefore, because of the minor role it plays and the lack of data, the lack of associations in the contingency plot, co-oncoplot, co-lollipop, and KM-survival plot makes sense. 